Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2020-04-06 review-responses.tex

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript, 
``Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes''.

In response to reviewer comments (and feedback from other sources)
we have substantially rewritten the introductory portions of the manuscript
and reworked the figures,
hopefully making it more accessible to a wider audience.
We think the manuscript is improved, and hope you agree.

@@

\pagebreak
\setcounter{page}{1}

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
It is most important that you address the following in a revised manuscript: thorough revisions of Figures 1 and 2, and the text in lines 80-101.
\end{quote}

We have thoroughly revised both figures, split Figure 1 into two figures,
and added an additional figure,
to ease the conceptual burden,
as well as reorganized the introductory section,
adding a more gentle introduction to the concepts.

@@

\begin{quote}
Ralph et al.\ describe a duality between summary statistics obtained from marginal trees in ancestral recombination graphs (ARG) and site-based summary statistics.
They use simulated data to demonstrate this leads to computational speed-ups and apply their methods using the ARG of 1000 Genomes data recently inferred by Speidel et al.
As the authors point out, the idea of a duality between genotypic and genealogical data is not new and has been extensively exploited in the past both theoretically and in practice,
but this paper presents a unified view that will likely be of interest to the readers of Genetics.
\end{quote}

Thanks -- we're glad you think it will be useful!

\begin{point}{Figure \ref{fig:divergence_diagram}}
    This is an important figure in the paper. I think the figure (and/or the legend) can be improved, mainly by adding a definition of all quantities being used (at the beginning of the legend or in a box within the figure).
    \begin{itemize}
        \item when first reading the paper, it takes some effort to understand what ``a'' and ``b'' represent in $f(a,b)$
            (and it doesn't help that this is a multi-panel figure with panels named the same way).
            It would also be good to make clear what the numbers ``3'' and ``2'' reflect, which requires some interpretation (this could be just stated explicitly, as for other parameters).
        \item ``seven mutations'', I believe the authors mean ``five mutations''.
        \item ``plus the weight of the node itself, if it is a sample'', this does not apply to this example and is confusing, it should probably be mentioned in Def 1 but not Fig 1.
        \item ``remaining weight of each mutation'' has not been defined in the figure, which is referred to before Def 1 is encountered.
    \end{itemize}
\end{point}

\reply{
    We've substantially reworked this figure,
    splitting out part (a) into \autoref{fig:adding_weights},
    simplifying it and providing more description in words.
    We have also added another figure before it
    to reinforce the same basic intuition.
    We've also rephrased the undefined phrase ``remaining weight''.
}

@@

\begin{point}{}
Figure 1 is not understandable. It's not even understandable to me now, having read the entire paper. Some feedback: I'm seeing trees with 5 samples, which are gray in the first, and blue/red in the second tree. I'm seeing five mutations as yellow stars (do the mutations relate to the red/blue?). I'm reading in the legend something about 7 mutations and 10 sites, and neither tof these two numbers are found somewhere in the figure. I see bracketed pairs of numbers next to nodes that I don't understand, and I see confusing equations. And the legend doesn't help.

I absolutely share a sense with the authors that there should be such a figure, but I think the choice of what information to show in the graphic is simply not working (for me). I don't really know what to do about it, since - as written above - I don't even understand why ``Node statistics'' need to be introduced at that point in the text. I have two comments that might help improving or replacing that figure:
    \begin{itemize}
        \item Consider using a different summary function. The sequence divergence is among the more complex statistics, which require two groups of samples. Why not taking something much simpler, such as pairwise nucleotide diversity?
        \item Consider replacing this figure with one that explains how Site statistics are calculated using weights and a summary function.
    \end{itemize}

I don't know, but at the very least, if you want to keep that figure more or less as is, you need to spend an entire paragraph explaining it properly to the reader. What's written in main text and legend about that figure is simply not sufficient!
\end{point}

\reply{
    Apologies that this figure was so confusing!
    We have substantially simplified this figure,
    split out \autoref{fig:adding_weights} into a separate figure,
    rewritten the explanation,
    and begun with a different, simpler figure.
    Hopefully it is more clear now!
    A different summary function is a good suggestion, but we actually think that divergence
    is easier to understand than diversity, as it doesn't require sampling without replacement.
}

### 2020-04-07 treestats_paper.tex

The requirement that $f(0) = f(\tiw) = 0$ ensures
that statistics do not depend on portions of the tree either not ancestral to any of the samples
or ancestral to all of them.
This is desireable because genetic variation within the samples
cannot inform us about such parts of the tree.
However, as we will see below it is sometimes useful to use non-strict summary functions.
% Below, in section XXX, we describe how to efficiently
% maintain a list of the weights of the subtrees inherit from each node
% as we iterate through the tree sequence.

@@

Under a neutral, infinite-sites model of mutation with constant mutation rate across time,
the expected number of mutations per branch is proportional to its length.
This implies an isormorphism between ``Site'' and ``Branch'' statistics defined above,
which is discussed in more detail in \citet{ralph2019empirical}.
For instance, the site statistic of Example \ref{ex:site_diversity} (genetic diversity)
and the branch statistic of Example \ref{ex:branch_diversity} (mean TMRCA)
use the same summary function $f(x) = x(n-x)/\left(n(n-1)\right)$.
These are closely related because under an infinite-sites model of mutation,
two sequences differ at a site only if there has been a mutation somewhere on the branches going back
to their most recent common ancestor.
Therefore, if mutations occur with constant rate
the expected value of genetic diversity
% averaging over mutational noise given the tree sequence,
is equal to the mutation rate multiplied by the average path distance between
the two samples in the trees.

>>>>

The requirement that $f(0) = f(\tiw) = 0$ ensures
that statistics do not depend on portions of the tree either not ancestral to any of the samples
or ancestral to all of them.
This is desirable because genetic variation within the samples
cannot inform us about such parts of the tree.
However, as we will see below it is sometimes useful to use non-strict summary functions.
% Below, in section XXX, we describe how to efficiently
% maintain a list of the weights of the subtrees inherit from each node
% as we iterate through the tree sequence.

@@

Under a neutral, infinite-sites model of mutation with constant mutation rate across time,
the expected number of mutations per branch is proportional to its length.
This implies an isomorphism between ``Site'' and ``Branch'' statistics defined above,
which is discussed in more detail in \citet{ralph2019empirical}.
For instance, the site statistic of Example \ref{ex:site_diversity} (genetic diversity)
and the branch statistic of Example \ref{ex:branch_diversity} (mean TMRCA)
use the same summary function $f(x) = x(n-x)/\left(n(n-1)\right)$.
These are closely related because under an infinite-sites model of mutation,
two sequences differ at a site only if there has been a mutation somewhere on the branches going back
to their most recent common ancestor.
Therefore, if mutations occur with constant rate
the expected value of genetic diversity
% averaging over mutational noise given the tree sequence,
is equal to the mutation rate multiplied by the average path distance between
the two samples in the trees.
